---
title: "Spectral parameterization for studying neurodevelopment: How and why"
subtitle: "Setting up Python in R Studio using the *reticulate* package"
author: "Brendan Ostlund, Thomas Donoghue, Berenice Anaya, Kelley E. Gunther, Sarah L. Karalunas, Bradley Voytek, & Koraly E. Pérez-Edgar"
output: html_document
---
This R Markdown file corresponds to data analysis for a **Developmental Cognitive Neuroscience** manuscript submitted as part of a special issue on EEG methods in developmental science in May, 2021. Participating families were a part of a study conducted by the [Cognition, Affect, and Temperament (CAT) lab, under the supervision of Koraly Pérez-Edgar](http://www.catlabpsu.com/) at Pennsylvania State University. The spectral parameterization ([specparam](https://fooof-tools.github.io/fooof/index.html)) algorithm was created by [Thomas Donoghue, Bradley Voytek, and colleagues from the Voytek lab](https://voyteklab.com/) at UC San Diego. Example data is available at the [Open Science Framework](https://osf.io/efj6c/) and GitHub pages for this project.

This tutorial shows how to access Python in R Studio via the *reticulate* package (Ushey et al., 2020). While there are multiple ways in which a user may call Python from R (e.g., **repl_python()** function), our discussion focuses solely on how to access Python in an R Markdown file. Please see the [*reticulate* website](https://rstudio.github.io/reticulate) for details on alternative approaches. We also recommend reviewing the [reticulate cheat sheet](https://ugoproto.github.io/ugo_r_doc/pdf/reticulate.pdf) provided by R Studio.

## **1.** Install packages
### Load reticulate 
```{r}
#install.packages("reticulate")
library(reticulate)
```

### Install miniconda
*Miniconda* is a small version of conda that includes python and common dependencies (https://docs.conda.io/projects/conda/en/latest/user-guide/install/download.html#anaconda-or-miniconda).
```{r}
#install_miniconda()
```

### Identify python version on your PATH
"It is not uncommon for several version of Python (and several conda or virtualenv environments within a given version) to be available on a given system. The reticulate package can bind to any of these versions, and in all cases will attempt to locate a version which includes the first Python package imported via the import() function."
```{r}
Sys.which("python")
```

## **2.** Identify and set python 
### Return all detected python versions
```{r}
py_discover_config()
```

### List all available conda environments
```{r}
conda_list(conda= "auto")
```

### Set python version 
This code chunk opens a .Rprofile file where the path to the preferred python version can be set. To do so, change the *PATH* listed in the **Sys.setenv(RETICULATE_PYTHON = *PATH*)** command to match the location where miniconda is installed. You will need to copy and paste the python path listed next to the "r-reticulate" option printed from the **conda_list(conda= "auto")** command from the code chunk above. R will need to be restarted so that the changes take effect. *You will only need to set this path once*; this information will be saved for subsequent use. 
```{r}
#install.packages("usethis")
usethis::edit_r_profile()
```

### Set virtual environment
```{r}
use_virtualenv("r-reticulate")
```


## **3.** Load python packages
### Install packages
```{r}
py_install("pandas", envname= "r-reticulate")
py_install("numpy", envname= "r-reticulate")
py_install("fooof", envname= "r-reticulate")
py_install("matplotlib", envname= "r-reticulate")
py_install("tqdm", envname= "r-reticulate")
py_install("pytest", envname= "r-reticulate")
```




