---
title: "Spectral parameterization for studying neurodevelopment: How and why"
subtitle: "Setting up Python in R Studio using the *reticulate* package"
author: "Brendan Ostlund, Thomas Donoghue, Berenice Anaya, Kelley E. Gunther, Sarah L. Karalunas, Bradley Voytek, & Koraly E. Pérez-Edgar"
output: html_document
---
This R Markdown file corresponds to data analysis for a **Developmental Cognitive Neuroscience** manuscript submitted as part of a special issue on EEG methods in developmental science in May, 2021. Participating families were a part of a study conducted by the [Cognition, Affect, and Temperament (CAT) lab, under the supervision of Koraly Pérez-Edgar](http://www.catlabpsu.com/) at Pennsylvania State University. The spectral parameterization ([specparam](https://fooof-tools.github.io/fooof/index.html)) algorithm was created by [Thomas Donoghue, Bradley Voytek, and colleagues from the Voytek lab](https://voyteklab.com/) at UC San Diego. Example data is available at the [Open Science Framework](https://osf.io/efj6c/) and GitHub pages for this project.

This tutorial shows how to access Python in R Studio via the *reticulate* package (Ushey et al., 2020). While there are multiple ways in which a user may call Python from R (e.g., **repl_python()** function), our discussion focuses solely on how to access Python in an R Markdown file. Please see the [*reticulate* website](https://rstudio.github.io/reticulate) for details on alternative approaches. We also recommend reviewing the [reticulate cheat sheet](https://ugoproto.github.io/ugo_r_doc/pdf/reticulate.pdf) provided by R Studio.

## **1.** Install packages
### Load reticulate
```{r}
#install.packages("reticulate")
library(reticulate)
```

### Install miniconda
*Miniconda* is a small version of conda that includes python and common dependencies (https://docs.conda.io/projects/conda/en/latest/user-guide/install/download.html#anaconda-or-miniconda).
```{r}
#install_miniconda()
```

### Identify python version on your PATH
Several Python (and conda) versions may already exist on your computer. It is therefore essential that we identify available versions and set the preferred versions for all work going forward.
```{r}
Sys.which("python")
```

## **2.** Identify and set python
### Return all detected python versions
```{r}
py_discover_config()
```

### List all available conda environments
```{r}
conda_list(conda= "auto")
```

### Set python version
This code opens a .Rprofile file where the path to the preferred python version can be set. To do so, change the *PATH* listed in the **Sys.setenv(RETICULATE_PYTHON = "*PATH*")** command to match the location where miniconda is installed. You will need to copy and paste the python path listed next to the "r-reticulate" option printed from the **conda_list(conda= "auto")** command from the code chunk above. Make sure that the new path is in quotations. R will need to be restarted so that the changes take effect. *You will only need to set this path once*; this information will be saved for subsequent use. We recommend running this code in the console; users have run into issues when running it in the markdown code chunk.
```{r}
#install.packages("usethis")
usethis::edit_r_profile()
```

### Set virtual environment
```{r}
use_virtualenv("r-reticulate")
```


## **3.** Load python packages
### Install packages
Uncomment this code when you are ready to install python packages.
```{r}
#py_install("pandas", envname= "r-reticulate")
#py_install("numpy", envname= "r-reticulate")
#py_install("fooof", envname= "r-reticulate")
#py_install("matplotlib", envname= "r-reticulate")
```
