---
title: "Spectral parameterization for studying neurodevelopment: How and why"
subtitle: "Fitting individual power spectrum"
author: "Brendan Ostlund, Thomas Donoghue, Berenice Anaya, Kelley E. Gunther, Sarah L. Karalunas, Bradley Voytek, & Koraly E. Pérez-Edgar"
output: html_document
---
Participating families were a part of a study conducted by the [Cognition, Affect, and Temperament (CAT) lab, under the supervision of Koraly Pérez-Edgar](http://www.catlabpsu.com/) at Pennsylvania State University. The spectral parameterization ([specparam](https://fooof-tools.github.io/fooof/index.html)) algorithm was created by [Thomas Donoghue, Bradley Voytek, and colleagues from the Voytek lab](https://voyteklab.com/) at UC San Diego. Example data is available at the [GitHub](https://github.com/fooof-tools/DevelopmentalDemo) page for this project. Please check the Github page periodically for updates.

Results are saved out to a **Output** folder for further consideration. Please note: In each Python code chunk, `message=FALSE` is set to suppress the reticulate message.


### Load R packages
These packages are required for this tutorial. Install packages using the *install.packages()* function if needed.
```{r, message= FALSE}
library(tidyverse)   # Access a collection of packages for data management (e.g., dplyr, ggplot2, readr)
library(reticulate)  # Interface Python and R Studio
library(magick)      # Load and adjust .PNG files
library(gridExtra)   # Arrange multiple plots
```

### Load Python libraries
Load the Python modules we need for this tutorial.
```{python, message=FALSE}
# Import some useful standard library modules
import os
from pathlib import Path

# Import some general scientific python libraries
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Import the parameterization model object
from fooof import FOOOF

# Import functions to visualize flattened power spectrum
from fooof.plts.spectra import plot_spectrum
from fooof.sim.gen import gen_aperiodic

# Import functions to examine frequency-by-frequency error of model fits
from fooof.analysis.error import compute_pointwise_error_fm
```

### Check specparam version
```{python, message=FALSE}
import fooof
print(fooof.__version__)
```

### Load individual PSD
Load CSV files, including:
- `freqs.csv`, which contains a vector of frequencies
- `eopPSDs.csv`, which contains the power values for an individual power spectrum
```{python, message=FALSE}
# Load csv files containing frequency and power values
freqs = np.ravel(pd.read_csv("../Data/freqs.csv"))
spectrum = np.ravel(pd.read_csv("../Data/indv.csv"))[1:101]
```

```{python, message=FALSE}
# Check shape of loaded data
print(freqs.shape)
print(spectrum.shape)
```

### Parameterize a power spectrum
Now we can parameterize our power spectrum!
```{python, message=FALSE}
# Initialize a model object for spectral parameterization, with some settings
fm = FOOOF(peak_width_limits = [2, 8], max_n_peaks = 6, min_peak_height = 0.05, verbose = False)
```

```{python, message=FALSE, fig.height= 6, fig.width= 9}
# Fit individual PSD over 3-40 Hz range
fm.report(freqs, spectrum, [3, 40])
plt.show()
```

```{python, message=FALSE}
# Save out a copy of the model fit report
fm.save_report('INDV_demo', file_path= '../Output')

# The following line can also be used to save out the model plot
#fm.plot(save_fig = True, file_name = 'INDV_demo', file_path = '../Output')
```

### Access model fit information
All of the model fit information is saved to the model object. Note that all attributes learned in the model fit process have a trailing underscore.
```{python, message=FALSE}
# Access the model fit parameters from the model object
print('Aperiodic parameters: \n', fm.aperiodic_params_, '\n')
print('Peak parameters: \n', fm.peak_params_, '\n')
print('Goodness of fit:')
print('Error - ', fm.error_)
print('R^2   - ', fm.r_squared_, '\n')
print('Number of fit peaks: \n', fm.n_peaks_)
```


### Extract periodic and aperiodic parameters
Another way to access model fit parameters is to use the `get_params` method, which can be used to access model fit attributes.
```{python, message=FALSE}
# Extract aperiodic and periodic parameter information
aps = fm.get_params('aperiodic_params')
peaks = fm.get_params('peak_params')
```

```{python, message=FALSE}
# Extract goodness of fit information
err = fm.get_params('error')
r2s = fm.get_params('r_squared')
```

```{python, message=FALSE}
# Extract specific parameters 
exp = fm.get_params('aperiodic_params', 'exponent')
cfs = fm.get_params('peak_params', 'CF')
```

```{python, message=FALSE}
# Print out a custom parameter report
template = ("With an error level of {error:1.2f}, specparam fit an exponent of {exponent:1.2f} and peaks of {cfs:s} Hz.")
print(template.format(error = err, exponent = exp, cfs = ' & '.join(map(str, [round(CF, 2) for CF in cfs]))))
```

### Plot flattened power spectrum
It may be useful to plot a flattened power spectrum, with the aperiodic fit removed. 
```{python, message=FALSE}
# Set whether to plot in log-log space
plt_log = False
```

```{python, message=FALSE}
# Do an initial aperiodic fit - a robust fit, that excludes outliers
init_ap_fit = gen_aperiodic(fm.freqs, fm._robust_ap_fit(fm.freqs, fm.power_spectrum))

# Recompute the flattened spectrum using the initial aperiodic fit
init_flat_spec = fm.power_spectrum - init_ap_fit
```

```{python, message=FALSE}
# Plot the flattened the power spectrum
plot_spectrum(fm.freqs, init_flat_spec, plt_log, label = 'Flattened spectrum', color = 'black')
plt.show()
```

## Move to R 
Transferring specparam model objects into R allows a researcher to work within an environment that they are comfortable navigating, while reducing the number of secondary programs needed to process data. Here, we provide code to call python objects into R using the `py$` command in a R code chunk. We also provide example code to visualize some of the plots included in the specparam module in ggplot, to allow the user additional control over plot dimensions. 

### Calling python objects into R
```{r}
# Transfer periodic parameters to R data frame
per <- as.data.frame(py$peaks) %>% 
         rename(CF= 1, PW= 2, BW= 3) %>% 
         mutate(peak_num= row_number()) %>% group_by(peak_num) %>% 
         mutate(index= seq_along(CF)) %>%
         pivot_wider(id_col= index, names_from= peak_num, values_from= c(CF,PW,BW)) %>% 
         select(index,CF_1,PW_1,BW_1,CF_2,PW_2,BW_2) 
```

```{r}
# Transfer aperiodic parameters to R data frame
aps <- as.data.frame(py$aps) %>% mutate(var= c("offset","exponent")) %>% spread(var,`py$aps`) %>% mutate(index= row_number()) %>% select(index,offset,exponent)
```

```{r}
# Transfer group fit information to R data frame
r2s <- as.data.frame(py$r2s) %>% rename(r2s= 1) %>% mutate(index= row_number())
err <- as.data.frame(py$err) %>% rename(err= 1) %>% mutate(index= row_number())
```

```{r}
# Pull IDs 
IDs <- read.csv("../Data/indv.csv", header= TRUE) %>% select(ID) %>% mutate(index= row_number())  

# Join data frames
dat <- full_join(IDs, per, by= "index") %>% 
         full_join(aps, by= "index") %>%
         full_join(r2s, by= "index") %>% 
         full_join(err, by= "index") %>% 
         select(-index) %>% arrange(ID)
```

### Save fit information
```{r}
# Save out data frame as csv file
write.csv(dat, "../Output/INDV_demo.csv", row.names= FALSE)
```

### Frequency-by-frequency error
It can be useful to plot frequency-by-frequency error of the model fit, to identify where in frequency space the spectrum is (or is not) being fit well. When fitting individual spectrum, this can be accomplished using the `compute_pointwise_error_fm` function.

In this case, we can see that error fluctuates around 0.05, which is the same as the mean absolute error for the model (MAE). There are points in the spectrum where the model fit is somewhat poor, particularly < 4 Hz, ~6-9 Hz, and ~14 Hz. Further considerations may be necessary for this model fit.
```{python, message=FALSE, fig.height= 6, fig.width= 9}
# Plot frequency-by-frequency error 
compute_pointwise_error_fm(fm, plot_errors = True)
plt.show()
```

```{python, message=FALSE}
# Return the frequency-by-frequency errors
errs_fm = compute_pointwise_error_fm(fm, plot_errors= False, return_errors = True)
errs_fm
```

```{python, message=FALSE}
# Note that the average of this error is the same as the global error stored
print('Average freq-by-freq error:\t {:1.3f}'.format(np.mean(errs_fm)))
print('specparam model fit error: \t\t {:1.3f}'.format(fm.error_))
```

