---
title: "Spectral parameterization for studying neurodevelopment: How and why"
subtitle: "Fitting multiple power spectra"
author: "Brendan Ostlund, Thomas Donoghue, Berenice Anaya, Kelley E. Gunther, Sarah L. Karalunas, Bradley Voytek, & Koraly E. Pérez-Edgar"
output: html_document
---
Participating families were a part of a study conducted by the [Cognition, Affect, and Temperament (CAT) lab, under the supervision of Koraly Pérez-Edgar](http://www.catlabpsu.com/) at Pennsylvania State University. The spectral parameterization ([specparam](https://fooof-tools.github.io/fooof/index.html)) algorithm was created by [Thomas Donoghue, Bradley Voytek, and colleagues from the Voytek lab](https://voyteklab.com/) at UC San Diego. Example data is available at the [GitHub](https://github.com/fooof-tools/DevelopmentalDemo) page for this project. Please check the Github page periodically for updates.

Results are saved out to a **Output** folder for further consideration. Please note: In each Python code chunk, `message=FALSE` is set to suppress the reticulate message. 


### Load R packages
These packages are required for this tutorial. Install packages using the *install.packages()* function if needed.
```{r, message = FALSE}
library(tidyverse)   # Access a collection of packages for data management 
library(reticulate)  # Interface Python and R Studio
library(magick)      # Load and adjust .PNG files
library(gridExtra)   # Arrange multiple plots
library(psych)       # Toolbox for data analysis
```

### Load Python libraries
Load the Python modules we need for this tutorial.
```{python, message = FALSE}
# Import some useful standard library modules
import os
from pathlib import Path

# Import some general scientific python libraries
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Import the parameterization model group object
from fooof import FOOOF,FOOOFGroup

# Import useful parameterization related utilities and plot functions
from fooof.bands import Bands
from fooof.analysis import get_band_peak_fg
from fooof.utils import trim_spectrum
from fooof.data import FOOOFSettings
from fooof.plts import plot_spectrum
from fooof.plts.periodic import plot_peak_fits, plot_peak_params
from fooof.plts.aperiodic import plot_aperiodic_params, plot_aperiodic_fits

# Import functions to examine frequency-by-frequency error of model fits
from fooof.analysis.error import compute_pointwise_error_fg
```

### Check specparam version
```{python, message = FALSE}
import fooof
print(fooof.__version__)
```

### Load group PSDs
Load CSV files, including:
- `freqs.csv`, which contains a vector of frequencies
- `eopPSDs.csv`, which contains the power values for a group of power spectrum, one for each subject
```{python, message = FALSE}
# Load csv files containing frequency and power values
freqs = np.ravel(pd.read_csv("../Data/freqs.csv"))
spectra = np.array(pd.read_csv("../Data/eop.csv"))[:, 1:101]
```

```{python, message = FALSE}
# Check shape of loaded data
print(freqs.shape)
print(spectra.shape)
```

```{python, message = FALSE}
# Get the number of participants
n_subjs = spectra.shape[0]
print('There are {:d} participants'.format(n_subjs))
```

### Fit power spectra
```{python, message = FALSE}
# Define `peak_width_limit` setting
peak_width = [1, 8]

# Define `max_n_peaks` setting
n_peaks = 6

# Define `min_peak_height` setting
peak_height = 0.10

# Define frequency range
PSD_range = [3, 40]
```

Now we can parameterize our group of power spectra!
```{python, message = FALSE}
# Initialize a model object for spectral parameterization, with some settings
fg = FOOOFGroup(peak_width_limits = peak_width, max_n_peaks = n_peaks, min_peak_height = peak_height, verbose = False)
```

```{python, message = FALSE}
# Fit group PSD over 3-40 Hz range
fg.fit(freqs, spectra, PSD_range)
```

```{python, message = FALSE}
# Save group results and plots of fit parameters
fg.save_report('EOP_demo', file_path = '../Output')

# The following line can also be used to save out the group plot
#fg.plot(save_fig = True, file_name = 'EOP_demo', file_path = '../Output')
```

### Extract periodic and aperiodic parameters
As with the individual model objects, the `get_param` method can be used to access model fit attributes. In addition, here we will use a `Bands` object and the `get_band_peak_fg` function to organize fit peaks into canonical band ranges.
```{python, message = FALSE}
# Extract aperiodic and periodic parameters
aps = fg.get_params('aperiodic_params')
per = fg.get_params('peak_params')
```

```{python, message = FALSE}
# Extract group fit information
err = fg.get_params('error')
r2s = fg.get_params('r_squared')
```

```{python, message = FALSE}
# Check the average number of fit peaks, per model
print('Average number of fit peaks: ', np.mean(fg.n_peaks_))
```

```{python, message = FALSE}
# Define canonical frequency bands
bands = Bands({'theta' : [4, 8], 
               'alpha' : [8, 13], 
               'beta' : [13, 30]})
```

```{python, message = FALSE}
# Extract band-limited peaks information
thetas = get_band_peak_fg(fg, bands.theta)
alphas = get_band_peak_fg(fg, bands.alpha)
betas =  get_band_peak_fg(fg, bands.beta)
```

### Plot model parameters
The specparam module also has functions for plotting the fit parameters.
```{python, message=  FALSE, fig.height = 6, fig.width = 6}
# Plot the measured aperiodic parameters
plot_aperiodic_params(aps)
plt.show()
```

```{python, message = FALSE, fig.height = 7, fig.width = 16}
# Plot the parameters for peaks, split up by band
_, axes = plt.subplots(1, 3, figsize = (14, 12))
all_bands = [thetas, alphas, betas]
for ax, label, peaks in zip(np.ravel(axes), bands.labels, all_bands):
    plot_peak_params(peaks, ax=ax)
    ax.set_title(label + ' peaks', fontsize = 24)
plt.subplots_adjust(hspace = 0.4)
plt.show()
```

### Plot model components
We can also plot constructions of model components. In the following, we plot reconstructed alpha peaks and aperiodic components. In this case, we can see an exceptionally flat aperiodic fits in the plot on the right, suggesting a potential outlier in the data set that warrants further consideration.  
```{python, message = FALSE, fig.height = 6, fig.width = 14}
# Plot reconstructions of model components
_, axes = plt.subplots(1, 2, figsize = (14, 6))
plot_peak_fits(alphas, ax = axes[0])
plot_aperiodic_fits(aps, fg.freq_range, ax = axes[1])
plt.show()
```

## Tuning the specparam algorithm 
There are no strict guidelines about optimal parameters that will be appropriate across data sets and recording modalities. We suggest applying a data-driven approach to tune model fitting for optimal performance, while taking into account your expectations about periodic and aperiodic activity given the data, the question of interest, and prior findings. 

One option is to parameterize a subset of data to evaluate the appropriateness of model fit settings prior to fitting each power spectrum in the data set. Here, we test parameters on a randomly selected 10% of the data. 

#### Subsample spectra to compare between models
First, lets randomly sub-sample 10% of the power spectra that we will use to test model settings.
```{python, message = FALSE}
# Set random seed
np.random.seed(1)
```

```{python, message = FALSE}
# Define settings for subsampling a selection of power spectra
subsample_frac = 0.10
n_sample = int(n_subjs * subsample_frac)
```

```{python, message = FALSE}
# Select a random selection of spectra to explore
select = np.random.choice(n_subjs, int(n_subjs * subsample_frac), replace = False)
spectra_subsample = spectra[select, :]
```

#### Define settings for each model
Here, we define settings for two models to be fit and compared.  
```{python, message = FALSE}
# Define `peak_width_limit` for each model
m1_peak_width = [2, 5]
m2_peak_width = [1, 8]

# Define `max_n_peaks` for each model
m1_n_peaks = 4
m2_n_peaks = 6

# Define `min_peak_height` for each model
m1_peak_height = 0.05
m2_peak_height = 0.10
```

#### Set frequency ranges for each model
To sub-select frequency ranges, we will use the `trim_spectrum` function to extract frequency ranges of interest for each model.
```{python, message = FALSE}
# Define frequency range for each model
m1_PSD_range = [2, 20]
m2_PSD_range = [3, 40]

# Sub-select frequency ranges
m1_freq, m1_spectra = trim_spectrum(freqs, spectra_subsample, m1_PSD_range)
m2_freq, m2_spectra = trim_spectrum(freqs, spectra_subsample, m2_PSD_range)
```

#### Fit models, with different settings
```{python, message = FALSE}
# Fit a model object with model 1 settings
fg1 = FOOOFGroup(peak_width_limits = m1_peak_width, max_n_peaks = m1_n_peaks, min_peak_height = m1_peak_height)
fg1.fit(m1_freq, m1_spectra)

# Create individual reports for model 1 settings
for ind in range(len(fg1)):
  temp_model = fg1.get_fooof(ind, regenerate = True)
  temp_model.save_report(file_name = 'EOP_' + str(ind) + '_fm1_settings', file_path = '../Output')
  )
```

```{python, message = FALSE}
# Fit a model object with model 2 settings
fg2 = FOOOFGroup(peak_width_limits = m2_peak_width, max_n_peaks = m2_n_peaks, min_peak_height = m2_peak_height)
fg2.fit(m2_freq, m2_spectra)

# Create individual reports for the model 2 settings
for ind in range(len(fg2)):
  temp_model = fg2.get_fooof(ind, regenerate = True)
  temp_model.save_report(file_name = 'EOP_' + str(ind) + '_fm2_settings', file_path = '../Output')
  )
```

#### Other ways to manage settings
Another way to manage model settings is with the `FOOOFSettings` object. Here we will redefine group model objects (`FOOOFGroup`), again using different settings for each one. 
```{python, message = FALSE}
# Define settings for model 1
settings1 = FOOOFSettings(peak_width_limits= m1_peak_width, max_n_peaks = m1_n_peaks, 
                          min_peak_height = m1_peak_height, peak_threshold = 2.,
                          aperiodic_mode = 'fixed')

# Define settings for model 2
settings2 = FOOOFSettings(peak_width_limits = m2_peak_width, max_n_peaks = m2_n_peaks, 
                          min_peak_height = m2_peak_height, peak_threshold = 2.,
                          aperiodic_mode = 'fixed')
```

#### Fit models with group model object
Note that when fitting power spectra, you can specify a fit range to fit the model to, so you don't have to explicitly trim the spectra. Here we will refit the example data, specifying the fit range, and then save the group reports. 
```{python, message = FALSE}
# Fit group PSD over the 2-20 Hz and 3-40 Hz ranges, respectively
fg1.fit(freqs, spectra_subsample, m1_PSD_range)
fg2.fit(freqs, spectra_subsample, m2_PSD_range)
```

```{python, message = FALSE}
# Save subset results and plots of fit parameters, for further examination
fg1.save_report(file_name = 'EOP_' + 'fg1_settings', file_path = '../Output')
fg2.save_report(file_name = 'EOP_' + 'fg2_settings', file_path = '../Output')
```

### Check model fits
After selecting initial model fit settings, and fitting  power spectra from the full sample, it is often worthwhile to check the goodness of model fit parameters. Please note, the model objects below correspond to the model fit at the top of this script.
```{python, message = FALSE, fig.height = 6, fig.width = 14}
# Plot distributions of goodness of fit parameters
# This information is presented in the print_reports output as well
fig, (ax0, ax1) = plt.subplots(1, 2, figsize = (14, 6))

ax0.hist(r2s)
ax0.set_xlabel('Variance explained (R^2)', fontsize = 20)
ax0.set_ylabel('Count', size = 20)
ax0.tick_params(labelsize = 18)

ax1.hist(err)
ax1.set_xlabel('Mean absolute error (MAE)', fontsize = 20)
ax1.set_ylabel('Count', size = 20)
ax1.tick_params(labelsize = 18)

fig.tight_layout()
plt.show()
```

### Underfitting
```{python, message = FALSE}
# Extract all fits that are above some error threshold, for further examination.
underfit_error_threshold = 0.10
underfit_check = []
for ind, res in enumerate(fg):
    if res.error > underfit_error_threshold:
        underfit_check.append(fg.get_fooof(ind, regenerate = True))
```

```{python, message = FALSE}
# Loop through the problem fits and save out reports to check later.
for ind, fm in enumerate(underfit_check):
  fm.save_report('EOP_UnderFit_ToCheck_#' + str(ind), file_path = '../Output')
  )
```

### Overfitting
```{python, message = FALSE}
# Extract all fits that are below some error threshold, for further examination.
overfit_error_threshold = 0.02
overfit_check = []
for ind, res in enumerate(fg):
    if res.error < overfit_error_threshold:
        overfit_check.append(fg.get_fooof(ind, regenerate = True))
```

```{python, message = FALSE}
# Loop through the problem fits and save out reports to check later.
for ind, fm in enumerate(overfit_check):
  fm.save_report('EOP_OverFit_ToCheck_#' + str(ind), file_path = '../Output')
  )
```


## Move to R 
Transferring specparam model objects into R allows a researcher to work within an environment that they are comfortable navigating, while reducing the number of secondary programs needed to process data. Here, we provide code to call python objects into R using the `py$` command in a R code chunk. We also provide example code to visualize some of the plots included in the specparam module in ggplot, to allow the user additional control over plot dimensions. 

### Calling python objects into R
```{r}
# Transfer periodic parameters to R data frame
per <- as.data.frame(py$per) %>% 
         rename(CF = 1, PW = 2, BW = 3, index = 4) %>% 
         group_by(index) %>% 
         mutate(peak_num = seq_along(CF), index = index + 1) %>% 
         pivot_wider(id_col = index, names_from = peak_num, values_from = c(CF, PW, BW)) %>% 
         select(index,
                CF_1, PW_1, BW_1, CF_2, PW_2, BW_2, CF_3, PW_3, BW_3,
                CF_4, PW_4, BW_4, CF_5, PW_5, BW_5, CF_6, PW_6, BW_6) 
```

```{r}
# Transfer band-limited identified peaks to R data frame
thetas <- as.data.frame(py$thetas) %>% 
            rename(CF_theta = 1, PW_theta = 2, BW_theta = 3) %>%
            mutate_at(vars(1:3), ~ifelse(. == "NaN", NA, .)) %>% 
            mutate(index = row_number())

alphas <- as.data.frame(py$alphas) %>% 
            rename(CF_alpha = 1, PW_alpha = 2, BW_alpha = 3) %>% 
            mutate_at(vars(1:3), ~ifelse(. == "NaN", NA, .)) %>% 
            mutate(index = row_number())

betas  <- as.data.frame(py$betas) %>% 
            rename(CF_beta = 1, PW_beta = 2, BW_beta = 3) %>% 
            mutate_at(vars(1:3), ~ifelse(. == "NaN", NA, .)) %>% 
            mutate(index = row_number())
```

```{r}
# Transfer aperiodic parameters to R data frame
aps <- as.data.frame(py$aps) %>% 
         rename(offset = 1, exponent = 2) %>% 
         mutate(index = row_number())
```

```{r}
# Transfer group fit information to R data frame
r2s <- as.data.frame(py$r2s) %>% 
         rename(r2s = 1) %>% 
         mutate(index = row_number())

err <- as.data.frame(py$err) %>% 
         rename(err = 1) %>% 
         mutate(index = row_number())
```

```{r}
# Read in IDs 
IDs <- read.csv("../Data/eop.csv", header = TRUE) %>% 
         select(ID) %>% 
         mutate(index = row_number())  

# Join data frames
dat <- full_join(IDs, thetas, by = "index") %>% 
         full_join(alphas, by = "index") %>%
         full_join(betas, by = "index") %>%
         full_join(aps, by = "index") %>%
         full_join(r2s, by = "index") %>% 
         full_join(err, by = "index") %>% 
         full_join(per, by = "index") %>% 
         select(-index) %>% arrange(ID)
```

### Save fit information
```{r}
# Save out data frame as CSV file
write.csv(dat, "../Output/EOP_demo.csv", row.names = FALSE)
```

```{r}
dat %>% select(-ID) %>% describe(skew = FALSE)
```

### Plot periodic and aperiodic parameters 
Here, we provide example code to recreate specparam narrowband plots using ggplot, for additional flexibility in visualizing periodic and aperiodic parameters.  
```{r, warning = FALSE, message = FALSE, fig.height = 4, fig.width = 16}
# Add a constant for group plots
dat <- dat %>% mutate(constant = 1)

# Plot offset
fig_a <- ggplot(dat, aes(x = constant, y = offset, fill = constant)) +
          geom_jitter(alpha = 0.7, width = 0.01, size = 2, aes(color = constant)) +
          theme_classic() +
          labs(title = "Offset") +
          stat_summary(fun = mean, shape = 23, size = 1.10, fill = "#D55E00") +
          scale_x_discrete(name = "", expand = c(1, 0), labels = NULL) +
          scale_y_continuous(name = "", 
                             limits = c(0.25, 2.25), breaks = seq(0.25, 2.25, 0.5), expand = c(0, 0)) +
          theme(axis.text = element_text(size = 12, color = "black"), 
                axis.title = element_text(size = 14), 
                plot.title = element_text(face = "bold")) +
          theme(legend.position = "none")

# Plot exponent
fig_b <- ggplot(dat, aes(x = constant, y = exponent, fill = constant)) +
          geom_jitter(alpha = 0.7, width = 0.01, size = 2, aes(color = constant)) +
          theme_classic() +
          labs(title = "Exponent") +
          stat_summary(fun = mean, shape = 23, size= 1.10, fill = "#D55E00") +
          scale_x_discrete(name = "", expand = c(1, 0), labels = NULL) +
          scale_y_continuous(name = "", 
                             limits = c(0.25, 2.25), breaks= seq(0.25, 2.25, 0.5), expand= c(0, 0)) +
          theme(axis.text = element_text(size = 12, color = "black"), 
                axis.title = element_text(size = 14), 
                plot.title = element_text(face = "bold")) +
          theme(legend.position = "none")

# Plot theta peaks
fig_c <- ggplot(dat, aes(x = CF_theta, y = PW_theta, size = BW_theta, fill = BW_theta)) +
          geom_point(alpha = 0.7, color = "#0072B2") +
          theme_classic() +
          labs(title = "Theta (4-8Hz)") +
          scale_x_continuous(name = "Center Frequency", 
                             limits= c(3.75, 8.25), breaks= seq(4, 8, 1), expand= c(0, 0)) +
          scale_y_continuous(name = "Power", 
                             limits = c(0, 1.35), breaks = seq(0, 1.25, 0.25), expand = c(0, 0)) +
          theme(axis.text = element_text(size = 12, color = "black"), 
                axis.title = element_text(size = 14), 
                plot.title = element_text(face = "bold")) +
          scale_size(range = c(4, 10)) +
          theme(legend.position = "none")

# Plot alpha peaks
fig_d <- ggplot(dat, aes(x = CF_alpha, y = PW_alpha, size = BW_alpha, fill = BW_alpha)) +
          geom_point(alpha = 0.7, color = "#0072B2") +
          theme_classic() +
          labs(title = "Alpha (8-13Hz)") +
          scale_x_continuous(name = "Peak Frequency", 
                             limits = c(7.75, 13.25), breaks = seq(8, 13, 1), expand = c(0, 0)) +
          scale_y_continuous(name = "Power", 
                             limits = c(0, 1.35), breaks = seq(0, 1.25, 0.25), expand = c(0, 0)) +
          theme(axis.text = element_text(size = 12, color = "black"), 
                axis.title = element_text(size = 14), 
                plot.title = element_text(face = "bold")) +
          scale_size(range = c(4, 10)) +
          theme(legend.position = "none")

# Plot beta peaks
fig_e <- ggplot(dat, aes(x = CF_beta, y = PW_beta, size = BW_beta, fill = BW_beta)) +
          geom_point(alpha = 0.7, color = "#0072B2") +
          theme_classic() +
          labs(title = "Beta (13-30Hz)") +
          scale_x_continuous(name = "Peak Frequency", 
                             limits = c(12.50, 30.50), breaks = seq(13, 30, 3), expand = c(0, 0)) +
          scale_y_continuous(name = "Power", 
                             limits = c(0, 1.35), breaks = seq(0, 1.25, 0.25), expand = c(0, 0)) +
          theme(axis.text = element_text(size = 12, color = "black"), 
                axis.title = element_text(size = 14), 
                plot.title = element_text(face = "bold")) +
          scale_size(range = c(4, 10)) +
          theme(legend.position = "none")

# Arrange plots
grid.arrange(fig_a, fig_b, fig_c, fig_d, fig_e, nrow = 1)
```

### Plot raw power spectra
Here, we provide code to plot the raw power spectra. In this case, we can transform data from wide to long format, and then plot individual (gray) and group averaged (blue) power across frequencies. 
```{r}
# Transform data from wide to long format
dat_long <- read.csv("../Data/eop.csv", header = TRUE) %>% 
              select(ID, X3:X40) %>%
              gather(Hz, power, X3:X40) %>% 
              arrange(ID) %>% 
              mutate(Hz = as.numeric(gsub('\\X', '', Hz))) 
```

```{r, fig.height = 5, fig.width = 5}
# Plot raw and mean power across frequencies
ggplot(dat_long, aes(x = Hz, y = power, group = ID)) +
  geom_line(color = "gray75") + 
  theme_classic() +
  stat_summary(aes(group = 1), geom = "line", fun = mean, size = 1.5, color = "#0072B2")  + 
  scale_x_continuous(name = "Frequency", 
                     limits = c(3, 40), breaks= seq(0, 40, 5), expand = c(0, 0)) +
  scale_y_log10(name = "Log(Power)") +
  theme(axis.text.x = element_text(size = 11, color = "black"), 
        axis.text.y = element_blank(), 
        axis.title = element_text(size = 14))
```

```{r}
# Print group fit results, if PNG file saved out
#print(image_read("../Output/EOP_demo.png"))
```

### Frequency-by-frequency error
It can be useful to plot frequency-by-frequency error of the model fit, to identify where in frequency space the spectrum is (or is not) being fit well. When fitting individual spectrum, this can be accomplished using the `compute_pointwise_error_fg` function. When plotting the error, the plot line is the mean error per frequency, across fits, and the shading indicates the standard deviation of the error, also per frequency.

In this case, we can see that error fluctuates around 0.03, which is the same as the mean absolute error for this group fit. There are points in the spectrum where the model fit is somewhat poor, particularly < 4 Hz. The code below lets you identify the highest mean error and largest standard deviation in error for the group fit. In this case, that occurs at 3 Hz, suggesting potential issues with fit at the lower end of the examined frequency range.
```{python, message = FALSE, fig.height = 6, fig.width = 9}
# Plot frequency-by-frequency error 
compute_pointwise_error_fg(fg, plot_errors = True)
plt.show()
```

```{python, message = FALSE}
# Return the errors - this returns a 2D matrix of errors for all fits
errs_fg = compute_pointwise_error_fg(fg, plot_errors = False, return_errors = True)
```

```{python, message = FALSE}
# Check which frequency has the highest error
f_max_err = fg.freqs[np.argmax(np.mean(errs_fg, 0))]
print('Frequency with highest mean error: \t\t\t', f_max_err)
```

```{python, message = FALSE}
# Check which frequency has the largest standard deviation of error
f_max_std = fg.freqs[np.argmax(np.std(errs_fg, 0))]
print('Frequency with highest standard deviation of error: \t', f_max_std)
```

In some cases, it may be necessary to drop poor (or failed) model fits from the group object. This can be done using the `fg.drop` function, as shown here. In this case, we remove a participant who has a MAE greater than 0.10. The error threshold will vary depending on sample characteristics and data quality.
```{python, message = FALSE}
# Drop poor model fits based on MAE
fg.drop(fg.get_params('error') > 0.10)
```

