---
title: "Spectral parameterization for studying neurodevelopment: How and why"
subtitle: "Illustrative example using the spectral parameterization (specparam) algorithm"
author: "Brendan Ostlund, Thomas Donoghue, Berenice Anaya, Kelley E. Gunther, Sarah L. Karalunas, Bradley Voytek, & Koraly E. Pérez-Edgar"
output: html_document
---
Participating families were a part of a study conducted by the [Cognition, Affect, and Temperament (CAT) lab, under the supervision of Koraly Pérez-Edgar](http://www.catlabpsu.com/) at Pennsylvania State University. The spectral parameterization ([specparam](https://fooof-tools.github.io/fooof/index.html)) algorithm was created by [Thomas Donoghue, Bradley Voytek, and colleagues from the Voytek lab](https://voyteklab.com/) at UC San Diego. Example data is available at the [GitHub](https://github.com/fooof-tools/DevelopmentalDemo) page for this project. Please check the Github page periodically for updates.

Analyses for the illustrative example are based on the eyes-closed condition of a resting baseline task (*n* = 60). Please see the corresponding manuscript for more details. A few code chunks described the *03-R_GroupPSDs.Rmd* tutorial have been merged in the script below to reduce length. Demographic information was reported in the manuscript but are not examined here in order to protect confidentiality of our participants. Results are saved out to a **Output** folder for further consideration.

### Load R libraries
```{r, message= FALSE}
library(tidyverse)   # Access a collection of packages for data management (e.g., dplyr, ggplot2, readr)
library(reticulate)  # Interface Python and R Studio
library(magick)      # Load and adjust .PNG files
library(gridExtra)   # Arrange multiple plots
library(psych)       # Toolbox for data analysis
```

### Load Python libraries
```{python, message= FALSE}
# Import some useful standard library modules
import os
from pathlib import Path

# Import some general scientific python libraries
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Import the parameterization model group object
from fooof import FOOOF, FOOOFGroup

# Import useful parameterization related utilities and plot functions
from fooof.bands import Bands
from fooof.analysis import get_band_peak_fg 
from fooof.plts.periodic import plot_peak_fits, plot_peak_params
from fooof.plts.aperiodic import plot_aperiodic_params, plot_aperiodic_fits

# Import functions to examine frequency-by-frequency error of model fits
from fooof.analysis.error import compute_pointwise_error_fg
```


## Tuning the specparam algorithm 
### Load group PSDs
```{python, message=FALSE}
freqs = np.ravel(pd.read_csv("../Data/freqs.csv"))
spectra = np.array(pd.read_csv("../Data/ecl.csv"))[:,1:101]

print(freqs.shape)
print(spectra.shape)
```

### Test model fit settings on subset (10%) of data 
```{python, message=FALSE}
spectra_sub = pd.DataFrame(spectra)
spectra_sub = spectra_sub.sample(frac = 0.10, random_state = 2)

spectra_sub_array = spectra_sub.to_numpy()
```

```{python, message=FALSE}
# Create frequency vector ranging from 2-20 Hz
# NOTE: Python uses zero indexing
freq1 = freqs[2:39]

for index, row in spectra_sub.iterrows():
  spectrum = spectra_sub.loc[[index]].transpose()[2:39]      
  spectrum = spectrum.to_numpy()
  spectrum = np.ravel(spectrum)
  
  fm = FOOOF(peak_width_limits = [2, 5], max_n_peaks = 4, min_peak_height = 0.05)
  fm.fit(freq1, spectrum, [2,20])
  
  fm.save_report(file_name= 'ECL_' + str(index) + '_fm1_settings', file_path= '../Output')
  )
```

```{python, message=FALSE}
# Create frequency vector ranging from 3-40 Hz
# NOTE: Python uses zero indexing
freq2 = freqs[4:79]

for index, row in spectra_sub.iterrows():
  spectrum = spectra_sub.loc[[index]].transpose()[4:79]     
  spectrum = spectrum.to_numpy()
  spectrum = np.ravel(spectrum)
  
  fm = FOOOF(peak_width_limits = [1, 8], max_n_peaks = 6, min_peak_height = 0.10)
  fm.fit(freq2, spectrum, [3,40])
  
  fm.save_report(file_name= 'ECL_' + str(index) + '_fm2_settings', file_path= '../Output')
  )
```

```{python, message=FALSE}
# Initialize model objects for spectral parameterization, with some settings
fg1= FOOOFGroup(peak_width_limits= [2, 5], max_n_peaks = 4, min_peak_height = 0.05)
fg2= FOOOFGroup(peak_width_limits= [1, 8], max_n_peaks = 6, min_peak_height = 0.10)
```

```{python, message=FALSE}
# Fit group PSD over the 2-20 Hz and 3-40 Hz ranges, respectively
fg1.fit(freqs, spectra_sub_array, [2, 20])
fg2.fit(freqs, spectra_sub_array, [3, 40])
```

```{python, message=FALSE}
# Print and save subset results and plots of fit parameters, for further examination
fg1.save_report(file_name= 'ECL_' + 'fg1_settings', file_path= '../Output')
fg2.save_report(file_name= 'ECL_' + 'fg2_settings', file_path= '../Output')
```

### Inspect group power spectra
Based on model fit results from the subset of participants, published findings, and our expectations of data in this sample, we will move forward with fitting the group power spectra over the 3-40 Hz frequency range with settings from the *fg2* model object.
```{python, message=FALSE}
# Initialize a model object for spectral parameterization, with some settings
fg= FOOOFGroup(peak_width_limits= [1, 8], max_n_peaks = 6, min_peak_height = 0.10)
```

```{python, message=FALSE}
# Fit group PSD over the 3-40 Hz range
fg.fit(freqs, spectra, [3, 40])
```

```{python, message=FALSE}
# Save out group results and plots of fit parameters
fg.save_report('ECL_example', file_path= '../Output')
```

### Extract parameters
```{python, message=FALSE}
# Extract aperiodic and periodic parameters
aps = fg.get_params('aperiodic_params')
per = fg.get_params('peak_params')

# Extract group fit information
err = fg.get_params('error')
r2s = fg.get_params('r_squared')
```

```{r}
# Read in IDs 
IDs <- read.csv("../Data/ecl.csv", header= TRUE) %>% select(ID) %>% mutate(index= row_number())  

# Transfer periodic parameters to data frame
per <- as.data.frame(py$per) %>% 
            rename(CF= 1, PW= 2, BW= 3, index= 4) %>% 
            group_by(index) %>% mutate(peak_num= seq_along(CF), index= index + 1) %>% 
            pivot_wider(id_col= index, names_from= peak_num, values_from= c(CF,PW,BW)) %>% 
            select(index,CF_1,PW_1,BW_1,CF_2,PW_2,BW_2,CF_3,PW_3,BW_3,CF_4,PW_4,BW_4,CF_5,PW_5,BW_5,CF_6,PW_6,BW_6)

# Transfer aperiodic parameters to data frame
aps <- as.data.frame(py$aps) %>% rename(offset= 1, exponent= 2) %>% mutate(index= row_number())

# Transfer group fit information to data frame
r2s <- as.data.frame(py$r2s) %>% rename(r2s= 1) %>% mutate(index= row_number())
err <- as.data.frame(py$err) %>% rename(err= 1) %>% mutate(index= row_number())

# Join data frames
dat <- full_join(IDs, per, by= "index") %>%
         full_join(aps, by= "index") %>% 
         full_join(r2s, by= "index") %>% 
         full_join(err, by= "index") %>% 
         select(-index) %>% arrange(ID)
```

### Check model fits
```{python, message=FALSE, fig.height= 6, fig.width= 14}
# Plot distributions of goodness of fit parameters
fig, (ax0, ax1) = plt.subplots(1, 2, figsize = (14, 6))

ax0.hist(r2s, density = True)
ax0.set_xlabel('Variance explained (R^2)', fontsize = 20)
ax0.set_ylabel('Count', size = 20)
ax0.tick_params(labelsize = 18)

ax1.hist(err, density = True)
ax1.set_xlabel('Mean absolute error (MAE)', fontsize = 20)
ax1.set_ylabel('Count', size = 20)
ax1.tick_params(labelsize = 18)

fig.tight_layout()
plt.show()
```

### Underfitting
```{python, message=FALSE}
# Extract all fits that are above some error threshold, for further examination.
underfit_error_threshold = 0.10
underfit_check = []
for ind, res in enumerate(fg):
    if res.error > underfit_error_threshold:
        underfit_check.append(fg.get_fooof(ind, regenerate=True))
```

There were no instances of MAE > 0.10 when fitting power spectra from the eyes-closed condition.
```{python, message=FALSE}
# Loop through the problem fits and saving out reports to check later.
for ind, fm in enumerate(underfit_check):
  fm.save_report('ECL_UnderFit_ToCheck_#' + str(ind), file_path= '../Output')
  )
```

### Overfitting
```{python, message=FALSE}
# Extract all fits that are below some error threshold, for further examination.
overfit_error_threshold = 0.02
overfit_check = []
for ind, res in enumerate(fg):
    if res.error < overfit_error_threshold:
        overfit_check.append(fg.get_fooof(ind, regenerate=True))
```

There was one instance identified as possibly being overfit (MAE < 0.02), ID# 204. After additional consideration, we decided that the identified peaks and aperiodic fit were reasonable for this participant. They were retained for subsequent analyses. 
```{python, message=FALSE}
# Loop through the problem fits and saving out reports to check later.
for ind, fm in enumerate(overfit_check):
  fm.save_report('ECL_OverFit_ToCheck_#' + str(ind), file_path= '../Output')
  )
```

### Frequency-by-frequency error
```{python, message=FALSE, fig.height= 6, fig.width= 9}
# Plot frequency-by-frequency error 
compute_pointwise_error_fg(fg, plot_errors = True)
plt.show()
```

```{python, message=FALSE}
# Return the errors - this returns a 2D matrix of errors for all fits
errs_fg = compute_pointwise_error_fg(fg, plot_errors = False, return_errors = True)

# Check which frequency has the highest error
f_max_err = fg.freqs[np.argmax(np.mean(errs_fg, 0))]
print('Frequency with highest mean error: \t\t\t', f_max_err)

# Check which frequency has the largest standard deviation of error
f_max_std = fg.freqs[np.argmax(np.std(errs_fg, 0))]
print('Frequency with highest standard deviation of error: \t', f_max_std)
```


## Primary analysis
For illustrative purposes, we examine differences among children who were identified as behaviorally inhibited (BI; n = 30) and non-behaviorally inhibited (BN; n = 30). Group membership in the behaviorally inhibited group was determined by a total score of greater than/equal to 119 and/or a social novelty score greater than/equal to 60.  Descriptive information for these children is provided in the accompanying manuscript. 

### Fit power spectra by group
```{python, message=FALSE}
spectra_BI = np.array(pd.read_csv("../Data/ecl_BI.csv"))[:, 1:101]
spectra_BN = np.array(pd.read_csv("../Data/ecl_BN.csv"))[:, 1:101]
```

```{python, message=FALSE}
# Initialize model objects for spectral parameterization
fg_BI= FOOOFGroup(peak_width_limits= [1, 8], max_n_peaks = 6, min_peak_height = 0.10, verbose = False)
fg_BN= FOOOFGroup(peak_width_limits= [1, 8], max_n_peaks = 6, min_peak_height = 0.10, verbose = False)
```

```{python, message=FALSE}
# Fit group PSD over the 3-40 Hz range
fg_BI.fit(freqs, spectra_BI, [3, 40])
fg_BN.fit(freqs, spectra_BN, [3, 40])
```

```{python, message=FALSE}
# Print and save subset results and plots of fit parameters, for further examination
fg_BI.save_report('specparam_ecl_BI', file_path= '../Output')
fg_BN.save_report('specparam_ecl_BN', file_path= '../Output')
```

### Plot components by group
```{python, message=FALSE}
# Set up labels and colors for plotting
colors = ['#E69F00', '#0072B2'] 
labels = ['BI', 'BN']
```

```{python, message=FALSE}
# Define frequency bands of interest
bands = Bands({'theta' : [4, 8],
               'alpha' : [8, 13],
               'beta' : [13, 30]})
```

```{python, message=FALSE, fig.height= 6, fig.width= 8}
# Extract theta peaks from each group
BI_thetas = get_band_peak_fg(fg_BI, bands.theta)
BN_thetas = get_band_peak_fg(fg_BN, bands.theta)
```

```{python, message=FALSE, fig.height= 6, fig.width= 8}
# Extract alpha peaks from each group
BI_alphas = get_band_peak_fg(fg_BI, bands.alpha)
BN_alphas = get_band_peak_fg(fg_BN, bands.alpha)

# Plot reconstructed alpha peaks
plot_peak_fits([BI_alphas, BN_alphas], colors = colors)
plt.show()
```

```{python, message=FALSE}
# Extract beta peaks from each group
BI_betas = get_band_peak_fg(fg_BI, bands.beta)
BN_betas = get_band_peak_fg(fg_BN, bands.beta)
```

```{python, message=FALSE, fig.height= 6, fig.width= 8}
# Extract the aperiodic parameters for each group
BI_aps = fg_BI.get_params('aperiodic_params')
BN_aps = fg_BN.get_params('aperiodic_params')

# Plot aperiodic fits
plot_aperiodic_fits([BI_aps, BN_aps], [3, 40], control_offset = True, log_freqs = False, labels = labels, colors = colors)
plt.show()
```

```{python, message=FALSE}
# Extract peak parameters for each group
BI_per = fg_BI.get_params('peak_params')
BN_per = fg_BN.get_params('peak_params')

# Extract group fit information for each group
BI_err = fg_BI.get_params('error')
BN_err = fg_BN.get_params('error')

BI_r2s = fg_BI.get_params('r_squared')
BN_r2s = fg_BN.get_params('r_squared')
```

### Call python objects into R
```{r}
# Transfer periodic parameters to R data frame
BI_per <- as.data.frame(py$BI_per) %>% 
            rename(CF= 1, PW= 2, BW= 3, index= 4) %>% 
            group_by(index) %>% mutate(peak_num= seq_along(CF), index= index + 1) %>% 
            pivot_wider(id_col= index, names_from= peak_num, values_from= c(CF,PW,BW)) %>% 
            select(index,CF_1,PW_1,BW_1,CF_2,PW_2,BW_2,CF_3,PW_3,BW_3,CF_4,PW_4,BW_4,CF_5,PW_5,BW_5)

BN_per <- as.data.frame(py$BN_per) %>% 
            rename(CF= 1, PW= 2, BW= 3, index= 4) %>% 
            group_by(index) %>% mutate(peak_num= seq_along(CF), index= index + 1) %>% 
            pivot_wider(id_col= index, names_from= peak_num, values_from= c(CF,PW,BW)) %>% 
            select(index,CF_1,PW_1,BW_1,CF_2,PW_2,BW_2,CF_3,PW_3,BW_3,CF_4,PW_4,BW_4,CF_5,PW_5,BW_5,CF_6,PW_6,BW_6)
```

```{r}
# Transfer band-limited identified peaks to R data frame
BI_thetas <- as.data.frame(py$BI_thetas) %>% rename(CF_theta= 1, PW_theta= 2, BW_theta= 3) %>%
              mutate_at(vars(1:3), ~ifelse(.== "NaN", NA, .)) %>% mutate(index= row_number())
BI_alphas <- as.data.frame(py$BI_alphas) %>% rename(CF_alpha= 1, PW_alpha= 2, BW_alpha= 3) %>% 
              mutate_at(vars(1:3), ~ifelse(.== "NaN", NA, .)) %>% mutate(index= row_number())
BI_betas  <- as.data.frame(py$BI_betas) %>% rename(CF_beta= 1, PW_beta= 2, BW_beta= 3) %>% 
              mutate_at(vars(1:3), ~ifelse(.== "NaN", NA, .)) %>% mutate(index= row_number())

BN_thetas <- as.data.frame(py$BN_thetas) %>% rename(CF_theta= 1, PW_theta= 2, BW_theta= 3) %>% 
              mutate_at(vars(1:3), ~ifelse(.== "NaN", NA, .)) %>% mutate(index= row_number())
BN_alphas <- as.data.frame(py$BN_alphas) %>% rename(CF_alpha= 1, PW_alpha= 2, BW_alpha= 3) %>%
              mutate_at(vars(1:3), ~ifelse(.== "NaN", NA, .)) %>% mutate(index= row_number())
BN_betas  <- as.data.frame(py$BN_betas) %>% rename(CF_beta= 1, PW_beta= 2, BW_beta= 3) %>%
              mutate_at(vars(1:3), ~ifelse(.== "NaN", NA, .)) %>% mutate(index= row_number())
```

```{r}
# Transfer aperiodic parameters to R data frame
BI_aps <- as.data.frame(py$BI_aps) %>% rename(offset= 1, exponent= 2) %>% mutate(index= row_number())
BN_aps <- as.data.frame(py$BN_aps) %>% rename(offset= 1, exponent= 2) %>% mutate(index= row_number())
```

```{r}
# Transfer group fit information to R data frame
BI_r2s <- as.data.frame(py$BI_r2s) %>% rename(r2s= 1) %>% mutate(index= row_number())
BI_err <- as.data.frame(py$BI_err) %>% rename(err= 1) %>% mutate(index= row_number())

BN_r2s <- as.data.frame(py$BN_r2s) %>% rename(r2s= 1) %>% mutate(index= row_number())
BN_err <- as.data.frame(py$BN_err) %>% rename(err= 1) %>% mutate(index= row_number())
```

```{r}
# Read in ID variable for each group
BI_IDs <- read.csv("../Data/ecl_BI.csv", header= TRUE) %>% select(ID) %>% mutate(GRP= "BI", index= row_number())  
BN_IDs <- read.csv("../Data/ecl_BN.csv", header= TRUE) %>% select(ID) %>% mutate(GRP= "BN", index= row_number())  
```

```{r}
# Join data frames for each group
dat_BI <- full_join(BI_IDs, BI_thetas, by= "index") %>% 
            full_join(BI_alphas, by= "index") %>%
            full_join(BI_betas, by= "index") %>%
            full_join(BI_aps, by= "index") %>%
            full_join(BI_r2s, by= "index") %>% 
            full_join(BI_err, by= "index") %>% 
            full_join(BI_per, by= "index") %>% 
            select(-index)

dat_BN <- full_join(BN_IDs, BN_thetas, by= "index") %>% 
            full_join(BN_alphas, by= "index") %>%
            full_join(BN_betas, by= "index") %>%
            full_join(BN_aps, by= "index") %>% 
            full_join(BN_r2s, by= "index") %>% 
            full_join(BN_err, by= "index") %>% 
            full_join(BN_per, by= "index") %>% 
            select(-index)

# Join group data frames
dat <- full_join(dat_BI, dat_BN) %>% arrange(ID)
```

```{r}
# Save out data frame 
write.csv(dat, "../Output/BI_example.csv", row.names= FALSE)
```

### Descriptive information
```{r, warning= FALSE, message= FALSE}
dat %>% select(-ID) %>% filter(GRP== "BI") %>% describe(skew= FALSE)
dat %>% select(-ID) %>% filter(GRP== "BN") %>% describe(skew= FALSE)
```

### Group comparisons
```{r}
# Aperiodic parameters
t.test(offset~ GRP, data= dat, var.equal= TRUE)
t.test(exponent~ GRP, data= dat, var.equal= TRUE)

# Alpha (8-13 Hz) parameters
t.test(CF_alpha~ GRP, data= dat, var.equal= TRUE)
t.test(PW_alpha~ GRP, data= dat, var.equal= TRUE)
t.test(BW_alpha~ GRP, data= dat, var.equal= TRUE)

# Beta (13-30 Hz) parameters
t.test(CF_beta~ GRP, data= dat, var.equal= TRUE)
t.test(PW_beta~ GRP, data= dat, var.equal= TRUE)
t.test(BW_beta~ GRP, data= dat, var.equal= TRUE)

# Goodness of fit parameters
t.test(r2s~ GRP, data= dat, var.equal= TRUE)
t.test(err~ GRP, data= dat, var.equal= TRUE)
```

### Plot raw power spectra
```{r}
# Read in ID variable for each group
BI_IDs <- read.csv("../Data/ecl_BI.csv", header= TRUE) %>% select(ID) %>% mutate(GRP= "BI")  
BN_IDs <- read.csv("../Data/ecl_BN.csv", header= TRUE) %>% select(ID) %>% mutate(GRP= "BN") 

# Merge IDs into data frame
dat_IDs <- full_join(BI_IDs,BN_IDs)

# Transform data from wide to long format
dat_long <- read.csv("../Data/ecl.csv", header= TRUE) %>% 
              left_join(dat_IDs,., by= "ID") %>% 
              select(ID,GRP,X3:X40) %>%
              gather(Hz, power, X3:X40) %>% arrange(ID) %>% 
              mutate(Hz= as.numeric(gsub('\\X', '', Hz))) 
```

```{r, fig.height= 5, fig.width= 5}
colors <- c('#E69F00', '#0072B2')

# Plot raw and mean power across frequencies
ggplot(dat_long, aes(x= Hz, y= power, group= ID, color= GRP)) +
  geom_line(alpha= 0.35, aes(color= GRP)) +
  theme_classic() +
  scale_color_manual(values= colors) +
  stat_summary(aes(group= GRP), geom= "line", fun= mean, size= 1.5)  +
  scale_x_continuous(name= "Frequency", limits= c(3,40), breaks= seq(0,40, by= 5), expand= c(0,0)) +
  scale_y_log10(name= "Log(Power)") +
  theme(axis.text.x= element_text(size= 11, color= "black"), axis.text.y= element_blank(), axis.title= element_text(size= 14)) +
  theme(legend.position= "none")
```

### Plot periodic and aperiodic parameters 
```{r, warning= FALSE, message= FALSE, fig.height= 8, fig.width= 14}
colors <- c('#E69F00', '#0072B2')

# Plot offset
fig_a <- ggplot(dat, aes(x= GRP, y= offset, fill= GRP)) +
          geom_jitter(alpha= 0.7, width= 0.01, size= 2, aes(color= GRP)) +
          theme_classic() +
          scale_color_manual(values= colors) +
          labs(title= "Offset") +
          stat_summary(fun= mean, shape= 23, size= 1.10, fill= colors) +
          scale_x_discrete(name= "", expand= c(1, 0), labels= NULL) +
          scale_y_continuous(name= "", limits= c(0.5,2.5), breaks= seq(0.5,2.5,0.5), expand= c(0, 0)) +
          theme(axis.text= element_text(size= 12, color= "black"), axis.title= element_text(size= 14), plot.title= element_text(face= "bold")) +
          theme(legend.position= "none")

# Plot exponent
fig_b <- ggplot(dat, aes(x= GRP, y= exponent, fill= GRP)) +
          geom_jitter(alpha= 0.7, width= 0.01, size= 2, aes(color= GRP)) +
          theme_classic() +
          scale_color_manual(values= colors) +
          labs(title= "Exponent") +
          stat_summary(fun= mean, shape= 23, size= 1.10, fill= colors) +
          scale_x_discrete(name= "", expand= c(1, 0), labels= NULL) +
          scale_y_continuous(name= "", limits= c(0.5,2.5), breaks= seq(0.5,2.5,0.5), expand= c(0, 0)) +
          theme(axis.text= element_text(size= 12, color= "black"), axis.title= element_text(size= 14), plot.title= element_text(face= "bold")) +
          theme(legend.position= "none")

# Plot theta peaks
fig_c <- ggplot(dat, aes(x= CF_theta, y= PW_theta, size= BW_theta, fill= BW_theta, color= GRP)) +
          geom_point(alpha= 0.7) +
          theme_classic() +
          scale_color_manual(values= colors) +
          labs(title= "Theta (4-8Hz)") +
          scale_x_continuous(name= "Center Frequency",limits= c(3.75,8.25), breaks= seq(4,8, by= 1), expand= c(0,0)) +
          scale_y_continuous(name= "Power", limits= c(0,1.85), breaks= seq(0,1.75, by= 0.25), expand= c(0,0)) +
          theme(axis.text= element_text(size= 12, color= "black"), axis.title= element_text(size= 14), plot.title= element_text(face= "bold")) +
          scale_size(range = c(4, 10)) +
          geom_vline(xintercept= 6.65, linetype= "dashed", color= "#E69F00", size= 1.50) +
          geom_vline(xintercept= 7.06, linetype= "dashed", color= "#0072B2", size= 1.50) +
          theme(legend.position= "none")

# Plot alpha peaks
fig_d <- ggplot(dat, aes(x= CF_alpha, y= PW_alpha, size= BW_alpha, fill= BW_alpha, color= GRP)) +
          geom_point(alpha= 0.7) +
          theme_classic() +
          scale_color_manual(values= colors) +
          labs(title= "Alpha (8-13Hz)") +
          scale_x_continuous(name= "Peak Frequency", limits= c(7.75,13.25), breaks= seq(8,13, by= 1), expand= c(0,0)) +
          scale_y_continuous(name= "Power", limits= c(0,1.85), breaks= seq(0,1.75, by= 0.25), expand= c(0,0)) +
          theme(axis.text= element_text(size= 12, color= "black"), axis.title= element_text(size= 14), plot.title= element_text(face= "bold")) +
          scale_size(range = c(4, 10)) +
          geom_vline(xintercept= 10.03, linetype= "dashed", color= "#E69F00", size= 1.50) +
          geom_vline(xintercept= 9.50, linetype= "dashed", color= "#0072B2", size= 1.50) +
          theme(legend.position= "none")

# Plot beta peaks
fig_e <- ggplot(dat, aes(x= CF_beta, y= PW_beta, size= BW_beta, fill= BW_beta, color= GRP)) +
          geom_point(alpha= 0.7) +
          scale_color_manual(values= colors) +
          theme_classic() +
          labs(title= "Beta (13-30Hz)") +
          scale_x_continuous(name= "Peak Frequency", limits= c(12.50,30.50), breaks= seq(13,30, by= 3), expand= c(0,0)) +
          scale_y_continuous(name= "Power", limits= c(0,1.85), breaks= seq(0,1.75, by= 0.25), expand= c(0,0)) +
          theme(axis.text= element_text(size= 12, color= "black"), axis.title= element_text(size= 14), plot.title= element_text(face= "bold")) +
          scale_size(range = c(4, 10)) +
          geom_vline(xintercept= 18.09, linetype= "dashed", color= "#E69F00", size= 1.50) +
          geom_vline(xintercept= 17.43, linetype= "dashed", color= "#0072B2", size= 1.50) +
          theme(legend.position= "none")

# Arrange plots
grid.arrange(fig_c, fig_d, fig_e, fig_a, fig_b, nrow= 2)
```

## Absolute and relative power
```{r}
# Calculate power based on canonical bands
pwr <- read.csv("../Data/ecl.csv", header= TRUE) %>% select(ID,X4:X30) %>% 
         rowwise() %>% 
         mutate(total_power= mean(c(X4,X4.5,X5,X5.5,X6,X6.5,X7,X7.5,X8,X8.5,X9,X9.5,X10,
                                    X10.5,X11,X11.5,X12,X12.5,X13,X13.5,X14,X14.5,X15,X15.5,X16,X16.5,X17,X17.5,X18,X18.5,X19,X19.5,X20,
                                    X20.5,X21,X21.5,X22,X22.5,X23,X23.5,X24,X24.5,X25,X25.5,X26,X26.5,X27,X27.5,X28,X28.5,X29,X29.5,X30)), 
                theta_abs= mean(c(X4,X4.5,X5,X5.5,X6,X6.5,X7,X7.5,X8)), 
                alpha_abs= mean(c(X8,X8.5,X9,X9.5,X10,X10.5,X11,X11.5,X12,X12.5,X13)), 
                beta_abs= mean(c(X13,X13.5,X14,X14.5,X15,X15.5,X16,X16.5,X17,X17.5,X18,X18.5,X19,X19.5,X20,X20.5,X21,X21.5,X22,X22.5,X23,X23.5,X24,X24.5,X25,X25.5,X26,X26.5,X27,X27.5,X28,X28.5,X29,X29.5,X30))) %>% 
         mutate(theta_rel= theta_abs/total_power, alpha_rel= alpha_abs/total_power, beta_rel= beta_abs/total_power) %>% 
         select(-c(X4:X30))

dat <- left_join(dat, pwr, by= "ID")
```

### Group differences in relative power (traditional method)
```{r, warning= FALSE, fig.height= 6, fig.width= 12}
# Alpha power
t.test(alpha_rel~ GRP, data= dat, var.equal= TRUE)

a1 <- ggplot(dat, aes(x= GRP, y= alpha_rel, fill= GRP)) +
        geom_jitter(alpha= 0.7, width= 0.05, size= 2, aes(color= GRP)) +
        theme_classic() +
        scale_color_manual(values= colors) +
        labs(title= "Relative alpha power") +
        stat_summary(fun= mean, shape= 23, size= 1.5, fill= c('#E69F00', '#0072B2')) +
        scale_x_discrete(name= "", expand= c(1, 0), labels= c("BI","BN")) +
        scale_y_continuous(name= "", limits= c(0,5), breaks= seq(0,5,1), expand= c(0, 0)) +
        theme(axis.text= element_text(size= 12, color= "black"), axis.title= element_text(size= 14), plot.title= element_text(face= "bold")) +
        theme(legend.position= "none")

# Beta power
t.test(beta_rel~ GRP, data= dat, var.equal= TRUE)

a2 <- ggplot(dat, aes(x= GRP, y= beta_rel, fill= GRP)) +
        geom_jitter(alpha= 0.7, width= 0.05, size= 2, aes(color= GRP)) +
        theme_classic() +
        scale_color_manual(values= colors) +
        labs(title= "Relative beta power") +
        stat_summary(fun= mean, shape= 23, size= 1.5, fill= c('#E69F00', '#0072B2')) +
        scale_x_discrete(name= "", expand= c(1, 0), labels= c("BI","BN")) +
        scale_y_continuous(name= "", limits= c(0,1), breaks= seq(0,1,0.25), expand= c(0, 0)) +
        theme(axis.text= element_text(size= 12, color= "black"), axis.title= element_text(size= 14), plot.title= element_text(face= "bold")) +
        theme(legend.position= "none")

# Visualize group differences
grid.arrange(a1,a2, nrow= 1)
```

### BIQ (continuous) comparisons
```{r}
# Read in BIQ data
biq <- read.csv("../Data/biq.csv", header= TRUE) %>% filter(COND== "ECL") %>% select(-COND) 
```

```{r}
# Join with main data frame
dat <- left_join(dat, biq, by= c("ID","GRP"))
```

```{r}
# Correlation between total BIQ and aperiodic-adjusted alpha 
cor.test(dat$Total_BIQ, dat$CF_alpha)
cor.test(dat$Total_BIQ, dat$PW_alpha)

# Correlation between total BIQ and relative alpha power (traditional method)
cor.test(dat$Total_BIQ, dat$alpha_rel)
```

```{r, warning= FALSE, fig.height= 5, fig.width= 5}
# Association between BIQ scores and alpha center frequency (specparam)
reg1 <- lm(CF_alpha ~ Total_BIQ, data= dat)                      
int1 <- coefficients(reg1)[1]
slp1 <- coefficients(reg1)[2]

ggplot(dat, aes(x= Total_BIQ, y= CF_alpha, fill= GRP)) +
  geom_jitter(alpha= 0.7, width= 0.05, size= 2.5, aes(color= GRP)) +
  theme_classic() +
  scale_color_manual(values= colors) +
  geom_abline(intercept= int1, slope= slp1, color= "grey20", linetype= "dashed", size= 1.25) +
  scale_x_continuous(name= "Total BIQ score", limits= c(35,175), breaks= seq(35,175,15), expand= c(0, 0)) +
  scale_y_continuous(name= "Alpha center frequency", limits= c(7.75,13), breaks= seq(8,13,1), expand= c(0, 0)) +
  theme(axis.text= element_text(size= 11, color= "black"), axis.title= element_text(size= 13)) +
  theme(legend.position= "none")
```

```{r, warning= FALSE, fig.height= 5, fig.width= 5}
# Association between BIQ scores and alpha power (specparam)
reg2 <- lm(PW_alpha ~ Total_BIQ, data= dat)                      
int2 <- coefficients(reg2)[1]
slp2 <- coefficients(reg2)[2]

ggplot(dat, aes(x= Total_BIQ, y= PW_alpha, fill= GRP)) +
  geom_jitter(alpha= 0.7, width= 0.05, size= 2.5, aes(color= GRP)) +
  theme_classic() +
  scale_color_manual(values= colors) +
  geom_abline(intercept = int2, slope= slp2, color= "grey20", linetype= "dashed", size= 1.25) +
  scale_x_continuous(name= "Total BIQ score", limits= c(35,175), breaks= seq(35,175,15), expand= c(0, 0)) +
  scale_y_continuous(name= "Alpha power (specparam)", limits= c(0,2), breaks= seq(0,2,0.5), expand= c(0, 0)) +
  theme(axis.text= element_text(size= 11, color= "black"), axis.title= element_text(size= 13)) +
  theme(legend.position= "none")
```

```{r, warning= FALSE, fig.height= 5, fig.width= 5}
# Association between BIQ scores and relative alpha power (traditional method)
reg3 <- lm(alpha_rel ~ Total_BIQ, data= dat)
int3 <- coefficients(reg3)[1]
slp3 <- coefficients(reg3)[2]

ggplot(dat, aes(x= Total_BIQ, y= alpha_rel, fill= GRP)) +
  geom_jitter(alpha= 0.7, width= 0.05, size= 2.5, aes(color= GRP)) +
  theme_classic() +
  scale_color_manual(values= colors) +
  geom_abline(intercept= int3, slope= slp3, color= "grey20", linetype= "dashed", size= 1.25) +
  scale_x_continuous(name= "Total BIQ score", limits= c(35,175), breaks= seq(35,175,15), expand= c(0, 0)) +
  scale_y_continuous(name= "Relative alpha power", limits= c(0,5), breaks= seq(0,5,1), expand= c(0, 0)) +
  theme(axis.text= element_text(size= 11, color= "black"), axis.title= element_text(size= 13)) +
  theme(legend.position= "none")
```

### Frontal asymmetry analyses
```{python, message=FALSE}
# Load data
freq_asm = np.ravel(pd.read_csv("../Data/freqs.csv"))[4:79]
spectra_asm = np.array(pd.read_csv("../Data/asm.csv"))[:, 2:83]
```

```{python, message=FALSE}
# Fit group PSDs
fg_asm= FOOOFGroup(peak_width_limits= [1, 8], max_n_peaks = 6, min_peak_height = 0.10)
fg_asm.fit(freq_asm, spectra_asm, [3, 40])
```

```{python, message=FALSE}
# Extract alpha peaks
band = Bands({'alpha' : [8, 13]})
asm_alphas = get_band_peak_fg(fg_asm, band.alpha)
```

```{r}
# # Read in group variables
# GRPs <- read.csv("../Data/ecl.csv", header= TRUE) %>% select(ID,GRP) %>% mutate(GRP= recode_factor(GRP, BLN= "BI", NT= "BN"))
# 
# # Read in traditional frontal asymmetry variable
# asm_trad <- read.csv("../Data/asm.csv", header= TRUE) %>% 
#               select(ID,hem,X08:X13) %>% 
#               mutate_at(vars(X08:X13), log) %>% 
#               rowwise() %>% mutate(trad_pwr = mean(c(X08,X09,X10,X11,X12,X13))) %>% 
#               select(ID,hem,trad_pwr)
# 
# # Call alpha peaks from specparam object
# asm_alphas <- as.data.frame(py$asm_alphas) %>% rename(CF_alpha= 1, PW_alpha= 2, BW_alpha= 3) %>% mutate_at(vars(1:3), ~ifelse(.== "NaN", NA, .)) 
# 
# # Join data frames and calculate frontal asymmetry variables
# asm <- cbind(asm_trad,asm_alphas) %>% 
#         pivot_wider(names_from= hem, values_from= c(trad_pwr,CF_alpha,PW_alpha,BW_alpha)) %>% 
#         mutate(ASM_trad= trad_pwr_right - trad_pwr_left, ASM_specparam= PW_alpha_right - PW_alpha_left) %>% 
#         full_join(., GRPs, by= "ID") %>% select(ID,GRP,everything())
```

```{r}
# # Descriptive information by group
# asm %>% filter(GRP== "BI") %>% select(ASM_trad,ASM_specparam) %>% describe(skew= FALSE)
# asm %>% filter(GRP== "BN") %>% select(ASM_trad,ASM_specparam) %>% describe(skew= FALSE)
```

```{r}
# # Group comparison based on method
# t.test(ASM_trad~ GRP, data= asm, var.equal= TRUE)
# t.test(ASM_specparam~ GRP, data= asm, var.equal= TRUE)
```

```{r, warning= FALSE, fig.height= 5, fig.width= 6}
# # Correlation between methods
# cor.test(asm$ASM_trad, asm$ASM_specparam, method= "pearson")
# 
# # Visualize association between methods
# ggplot(asm, aes(x= ASM_trad, y= ASM_specparam, color= GRP)) +
#   geom_point() +
#   scale_color_manual(values= colors) +
#   theme_classic() +
#   scale_x_continuous(name= "Frontal asymmetry (traditional)", limits= c(-0.65,0.65), breaks= seq(-0.6,0.6,0.15), expand= c(0, 0)) +
#   scale_y_continuous(name= "Frontal asymmetry (specparam)", limits= c(-0.65,0.65), breaks= seq(-0.6,0.6,0.15), expand= c(0, 0)) +
#   annotate("text", x= 0.30, y= -0.35, label= "r(54) = -0.09, p = ns", fontface= "bold", size= 4) +
#   theme(axis.text= element_text(size= 12, color= "black"), axis.title= element_text(size= 14),legend.title= element_blank(), legend.text= element_text(size= 12)) 
```
